---
knitr: 
    opts_chunk: 
      code-fold: show
      results: hold
  
---

::::: {#obj-chap03}
:::: {.my-objectives}
::: {.my-objectives-header}
Chapter section list
:::

::: {.my-objectives-container}



:::
::::
:::::

# Overview {#sec-02}

```{r}
#| label: setup
#| results: hold
#| include: false

base::source(file = paste0(here::here(), "/R/helper.R"))

ggplot2::theme_set(ggplot2::theme_bw())
```


There are two main ways to creating a plotly object: 

1. by transforming a {**ggplot2**} object via `plotly::ggplotly()` into a plotly object or 
2. by directly initializing a plotly object with `plotly::plot_ly()`/`plotly::plot_geo()`/`plotly::plot_mapbox()`.

Both approaches have somewhat complementary strengths and weaknesses, so it can pay off to learn both approaches.

This introductory chapter outlines some over-arching concepts related to plotly in general. It also provides definitions for terminology used throughout the book and introduces some concepts useful for understanding the infrastructure behind any plotly object.

::: {.callout-note #nte-unnecessary-details}
###### Unnesseary details?

> Most of these details aren’t necessarily required to get started with plotly, but it will inevitably help you get ‘un-stuck’, write better code, and do more advanced things with plotly. (p.23)

Is it --- didactically speaking --- a good idea to start with not necessary details?
:::

## Intro to `plot_ly()`

Any graph made with the {**plotly**} R package is powered by the JavaScript library [plotly.js](https://github.com/plotly/plotly.js). The `plotly::plot_ly()` function provides a ‘direct’ interface to plotly.js.


### First `plot_ly()` examples

::: {.my-code-collection}
:::: {.my-code-collection-header}
::::: {.my-code-collection-icon}
:::::
:::::: {#exm-02-intro-plotly}
::::::

::::

::::{.my-code-collection-container}

::: {.panel-tabset}

###### Load data

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-intro-plotly}
: Load and inspect data
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: load-data
#| eval: false

utils::data(diamonds, package = "ggplot2")

base::set.seed(100)
diamonds_sample <- tibble::as_tibble(diamonds) |> 
  dplyr::slice_sample(n = 10000)

my_create_folder("data")
my_save_data_file("Chapter2", diamonds_sample, "diamonds_sample.rds")
```

::: {.callout-caution #cau-02-memory-problems}
###### Memory problem with big `diamonds` dataset

After several calls to `plotly::plot_ly()` or `plotly::ggplotly()` I ran out of memory and the graphs didn't show up in the browser from this point on.

I therefore reduced the dataset from `r nrow(ggplot2::diamonds)` to 10^3 rows.
:::



::::
:::::


###### First try

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-first-try}
: First try with warning message
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-02-first-try
#| fig-cap: "My first trial of a visualization with {*plotly*) worked but produced a (warning?) message."

diamonds_sample <- base::readRDS("data/Chapter2/diamonds_sample.rds")

# create a visualizations of the diamonds dataset
plotly::plot_ly(diamonds_sample, x = ~cut)
```

***

**To see the big difference between a static graph and the interactivity caused by {*plotly*} move the mouse cursor over the graphic!**

If we assign variable names (e.g., `cut`, `clarity`, etc.) to visual properties (e.g., `x`, `y`, `color`, etc.) within `plotly::plot_ly()`, as done in @fig-02-first-try, the program tries to find a sensible geometric representation of that information for us.

::: {.callout-warning #wrn-02-warning-messages}
###### What does the warning "No trace type specified" mean?

As you can see I got a warning message. This happened with all three offered examples. For the first and third visualization a trace of "histogram" seems appropriate, for the second one a "histogram2d".). This message is neither mentioned in the book (first edition) nor in the online version of the book.

An internet recherche uncovered [two different methods to get rid of these messages](https://github.com/plotly/plotly.R/issues/1202).

1. Add the correct type as argument. For example `plotly::plot_ly(diamonds, x = ~cut, type = "histogram")`
2. Write a small function:

``` R
suppressPlotlyMessage <- function(p) {
  suppressMessages(plotly_build(p))
}

suppressPlotlyMessage(plotly::plot_ly(diamonds, x = ~cut))
```

The first one is easier to use and it is the more general method too.
:::

:::::{.my-remark}
:::{.my-remark-header}
This difference between documentation and running code is annoying. It makes people completely new to {*plotly*} uncertain, especially as this is the first example in the tutorial!
:::

:::::




::::
:::::


###### Second try

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-second-try}
: Second try without warning messages
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-02-second-try
#| fig-cap: "Three examples of visualizing categorical data with `plotly::plot_ly()`: (top) mapping `cut` to `x` yields a bar chart, (middle) mapping `cut` & `clarity` to `x` & `y` yields a heatmap, and (bottom) mapping `cut` & `clarity` to `x` & `color` yields a dodged bar chart."
#| results: hold

diamonds_sample <- base::readRDS("data/Chapter2/diamonds_sample.rds")

# create three visualizations of the diamonds dataset
plotly::plot_ly(diamonds_sample, x = ~cut, type = "histogram")
plotly::plot_ly(diamonds_sample, x = ~cut, y = ~clarity, type = "histogram2d")
plotly::plot_ly(diamonds_sample, x = ~cut, color = ~clarity, colors = "Accent", type = "histogram") |> 
    plotly::config(doubleClickDelay = 1000)
```

***

Here I have added the correct type to the `plotly::plot_ly()` argument to get rid of the warning message.

In the bottom panel of @fig-02-second-try `color` is used to map each level of diamond clarity to a different color, then `colors` is used to specify the range of colors (which, in this case, the "Accent" color palette from the {**RColorBrewer**} package, but one can also supply custom color codes or a color palette function like `grDevices::colorRamp()`).

::: {.callout-note #nte-change-double-click-delay}
###### Double click delay time changed

In the last example I have changed the delay for double clicks. This setting determines the maximum delay between two consecutive clicks to be interpreted as a double-click, measured in milliseconds. By default, the timing is set to 300 ms (less than half a second). My code sets the double-click delay to 1000 milliseconds (1 second).

Try the following: Zoom into the graphics by holding the mouse button and dragging the cursor to the desired zoom rectangle. Then return to the original graphics by a relaxed double click. Try this for the first and for the third graphics. You will see that in the first graph you must provide a double click with a very short delay to return to the full view, whereas in the third graphics it is much easier.
:::




::::
:::::

###### Using pipe

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-using-pipe}
: Use pipe to modify the generic `plotly::plot_ly()` call
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-02-using-pipe
#| fig-cap: "Create graphics with the native pipe adding a specific layer"

diamonds_sample <- base::readRDS("data/Chapter2/diamonds_sample.rds")

diamonds_sample |> 
  plotly::plot_ly(x = ~cut, type = "histogram")  |> 
  plotly::layout(title = "My beautiful histogram")
```

***

(Almost) every function anticipates a plotly object as input to it’s first argument and returns a modified version of that plotly object. Furthermore, that modification is completely determined by the input values to the function (i.e., it doesn’t rely on any side-effects, unlike, for example, base R graphics). 

Here I have added with `plotly::layout()` to modify by adding parts to the `plotly::plot_ly()` function using the (native) pipe (`|>`) which was introduced into R with version 4.1.0, released May 2021. The base pipe operator allows you to direct the left-hand side object as the first argument into the function call on the right-hand side. This facilitates creating and reading code. (But note that there are [some differences](https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/) to the {**magrittr**} pipe used in the Plotly book.)

There is a huge amount of different arguments for the `plotly::layout()` function (See R [Figure Reference: layout](https://plotly.com/r/reference/layout/).
::::
:::::

:::

::::
:::::


The `plotly::plot_ly()` function has numerous arguments that are unique to the R package (e.g., `color`, `stroke`, `span`, `symbol`, `linetype`, etc) and make it easier to encode data variables (e.g., diamond clarity) as visual properties (e.g., `color`). By default, these arguments map values of a data variable to a visual range defined by the plural form of the argument.

#### GGally example

{**GGally**} extends {**ggplot2**} by adding several functions to reduce the complexity of combining geometric objects with transformed data. Some of these functions include a pairwise plot matrix, a two group pairwise plot matrix, a parallel coordinates plot, a survival plot, and several functions to plot networks.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-02-ggally-ggcoef}
: Visualizing the coefficient estimates and their standard errors 
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-02-ggally-ggcoef
#| fig-cap: "Zooming in on a coefficient plot generated from the `GGally::ggcoef()` function"
#| warning: false
#| eval: false

diamonds_sample <- base::readRDS("data/Chapter2/diamonds_sample.rds")
m <- stats::lm(
  log(price) ~ log(carat) + cut, 
  data = diamonds_sample)

gg <- GGally::ggcoef(m)
# dynamicTicks means generate new axis ticks on zoom
plotly::ggplotly(
  gg, 
  dynamicTicks = TRUE,
  doubleClickDelay = 1000
  ) |> 
  plotly::partial_bundle() |> 
  plotly::toWebGL()
```

***

@fig-02-ggally-ggcoef shows how injecting interactivity into this plot allows us to query exact values and zoom in on the most interesting regions. See [video](https://vimeo.com/322362701) how to do it.

::::
:::::

